---
title: "An Analysis of Obesity Prevalence and Its Socio-Economic Determinants in Scotland: Evidence from the 2008–2012 Health Surveys"
author: "2917216C, 2958694Y, 2963382T, 3136735L"
format:
  pdf:
    documentclass: article
    fontsize: 11pt
    titlepage: false
geometry: 
- left=0.5in
- right=0.5in
- top=0.9in 
- bottom=0.9in
editor: visual
---

## Introduction

### Background

Obesity is a growing public health concern worldwide and is associated with an increased risk of various chronic conditions, including cardiovascular diseases, diabetes, and certain cancers. Monitoring the prevalence of obesity within a population is important for identifying trends over time and informing public health policies.

The Scottish Health Survey is designed to assess the health of individuals living in Scotland and to monitor changes in health-related behaviours and outcomes. The dataset used in this analysis contains information collected between 2008 and 2012, including demographic, socio-economic, and lifestyle factors alongside an indicator of obesity.

The primary aim of this study is to examine whether the prevalence of obesity in Scotland has changed over the survey years. In addition, the analysis will explore whether obesity prevalence differs across age groups, gender, employment status, and lifestyle behaviours such as fruit and vegetable consumption.

### Variables

The dataset contains the following variables:

-   **AgeGroup:** The age range of the individual, grouped in categories

-   **Sex:** The gender of the individual (Male/Female)

-   **Employment:** Employment status, used as an indicator of socio-economic status

-   **Fruit:** Whether the individual meets the recommended daily fruit intake (Yes/No)

-   **Veg:** Whether the individual meets the recommended daily vegetable intake (Yes/No)

-   **Year:** The year in which the survey was conducted (2008-2012)

-   **Obese:** Indicator variable showing whether the individual is classified as obese.

In this analysis, Obese is treated as the response variable. The remaining variables (age group, sex, employment status, and dietary factors) are considered explanatory variables to investigate potential differences in obesity prevalence across different population subgroups.

## Exploratory data Analysis

```{r}
#| echo: false
#| warning: false
#| message: false
data<-read.csv("DAProject7.csv")
data<-read.csv("DAProject7.csv", stringsAsFactors = T)
library(ggplot2)
library(tidyverse)
library(tidymodels)
library(gt)
library(MASS)
library(patchwork)
library(moderndive)
library(sjPlot)
library(performance)
library(kableExtra)
library(ggplot2)
library(broom)
library(yardstick)
library(pROC)
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-proportions 
#| tbl-cap: Summary of Obesity Status in dataset
library(dplyr)
library(scales)
library(knitr)

data %>%
  count(Obese) %>%
  mutate(
    Proportion = round(n / sum(n), 3),
    Percentage = percent(n / sum(n))
  ) %>%
  kable(
    col.names = c("Obesity Status", "Count", "Proportion", "Percentage"),
    align = "c"
  ) %>%
    kable_styling(font_size = 8)
```

An explanatory data analysis was conducted to provide an initial overview of the obesity prevalence and its distribution within the dataset. @tbl-proportions summarises the proportion of obesity status, showing that roughly 29.6% are classified as obese, while 70.4% are not. This indicates that obesity affects a substantial proportion of the population

To explore patterns in obesity prevalence, the proportion of individuals classified as obese was examined across demographic and lifestyle factors.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-height: 6
#| fig-width: 6
#| fig-align: center
#| label: fig-obesity-dist
#| fig-cap: "Obesity Distribution across Demographic Factors"
#| fig-pos: H
library(dplyr)
library(tidyr)
library(ggplot2)

data_long <- data %>%
  mutate(across(c(Year, AgeGroup, Sex, Employment), factor)) %>%
  mutate(Employment = str_wrap(as.character(Employment), width = 12)) %>%
  pivot_longer(
    cols = c(Year, AgeGroup, Sex, Employment),
    names_to = "Variable",
    values_to = "Category"
  )%>%
  mutate(Variable = factor(Variable, levels = c("AgeGroup", "Sex", "Year", "Employment")))

ggplot(data_long, aes(x = Category, fill = Obese)) +
  geom_bar(position = "fill") +
  facet_wrap(~ Variable, scales = "free_x",ncol=2) +
  theme_minimal(base_size = 10) +
  labs(
    title = "Obesity Distribution Across Demographic Factors",
    x = NULL,
    y = "Proportion"
  ) +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1, size = 7),
    legend.position = "right", 
    strip.text = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

The proportion of individuals classified as obese shows slight variation over time, suggesting potential changes in obesity prevalence across survey years. Modest differences are also observed between males and females. In addition, obesity prevalence varies across age groups, and differences across employment categories suggest a potential influence of socio-economic factors.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-height: 3
#| fig-width: 5
#| fig-align: center
#| label: fig-obesity-diet
#| fig-cap: "Obesity by Diet Factors"
#| fig-pos: H
data %>%
  pivot_longer(cols = c(Veg, Fruit),
               names_to = "Diet",
               values_to = "Intake") %>%
  ggplot(aes(x = Intake, fill = Obese)) +
  geom_bar(position = "fill") +
  facet_wrap(~ Diet) +
  theme_minimal() +
  theme(legend.position = "right")+
  labs(title = "Obesity by Diet Factors",
       x = "Obese Indication (Yes/No)",
       y = "Proportion") + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

Obesity prevalence was also explored across dietary factors, including fruit and vegetable intake. The distributions appear similar, although slight differences may be observed between individuals meeting and not meeting the recommended intake.

The exploratory analysis suggests that obesity prevalence varies across different groups. In particular, differences are observed across age groups, and socio-economic and vegetable intake factors. These patterns indicate that obesity is likely influenced by multiple variables. A formal statistical analysis will be conducted to further investigate these relationships and determine their significance.

## Formal Analysis

### Logistic Regression Model

To investigate whether obesity prevalence in the Scottish population varies with factors such as year, demographic and lifestyle, we consider the nature of the response. Obesity status is classified as a binary reponse where if individual is obese $Y_i = 1$ and $Y_i = 0$ otherwise.

Given this binary structure, a logistic regression model is suitable for simulating the probability of an individual being classified as obese. Specifically we assume that,

$$
Y_i \sim \text{Bernoulli}(p_i).
$$

```{r}
#| echo: false
#| warning: false
#| message: false
#| results: hide
#q1 model
#response modeled with respect to year
model_year <- glm(Obese~Year, data = data, family = binomial)
summary(model_year)
```

### Model Result

To assess whether the obesity prevalence differ by year, a logistic regression model was fitted with year being the only explanatory variable included. The model takes the form:

$$
\log \left( \frac{p_i}{1 - p_i} \right)
= \beta_0 + \beta_1 \cdot \text{Year}_i
$$

where $p_i$ the probability that individual $i$ is obese and $\beta_1$ is the difference in log-odds of year variable.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-model-year
#| tbl-cap: "Odds ratios from logistic regression model of obesity prevalence by year"
#| out.width: "85%"
tab_model(model_year)
```

@tbl-model-year shows that the estimated odds ratio for year included 1. This indicates that there is no statistically significant association between year and obesity prevalence. In addition, the p-value (0.586) is larger than the significance level (0.05) providing no evidence to reject the null hypothesis that year has no effect. This suggest that year alone is not a meaningful predictor for obesity status and obesity prevalence does not differ significantly by year.

Next, for the second question of interest, after considering alternative models and balancing model fit with interpret ability, the following logistic regression model was selected: $$
\log\left(\frac{p_i}{1-p_i}\right)
=
\beta_0 + \beta_1 \text{Sex}_i + \pmb{\beta}_2 \text{AgeGroup}_i + \pmb{\beta_3} \text{Employment}_i
+ \beta_4 \text{Veg}_i.
$$ where

-   $p_i$ the probability that individual $i$ is obese;

-   $\beta_0$ is the intercept of the model, corresponding to the baseline group;

-   $\beta_1{\text{Sex}}_i$ is the difference in log-odds of obesity between males and females;

-   $\pmb\beta_2{\text{AgeGroup}_i}$ is the differences in log-odds of obesity between each age category and the reference age group where $\pmb{\beta}_2$ is a vector of coefficients for age group ;

-   $\pmb\beta_3{\text{Employment}}_i$ is the differences in log-odds of obesity between each employment category and the reference employment group where $\pmb\beta_3$ is a vector of coefficients for employement;

-   $\beta_4{\text{Veg}}_i$ is the difference in log-odds of obesity between individuals who do and do not meet the recommended vegetable intake.

```{r}
#| echo: false
#| warning: false
#| message: false
#| results: hide
#q2 models

#testing the binomial model with just the additive terms.
modelred <- glm(Obese~ Year + Sex + AgeGroup + Employment + Veg + Fruit, data = data, family = binomial)
summary(modelred)

#did backward elimination using step function to get the final model as:
stats::step(modelred, direction = "backward")
modelred1 <- glm (Obese ~ Sex + AgeGroup + Employment + Veg, data = data, family = binomial)

#Testing model with a few chosen interaction terms 
modelfull <-  glm(Obese ~ Year + Sex + AgeGroup + Employment + Veg + Fruit +
            Year:Sex + Year:AgeGroup + Sex:AgeGroup,
          data = data, family = binomial)
summary(modelfull)

#backward elimintation using step function again
stats::step(modelfull, direction = "backward")

# got to the final model for this as:
modelfull1 <-  glm(Obese ~ Sex + AgeGroup + Employment + Veg + Sex:AgeGroup,
                    data = data, family = binomial)

#tested the both model (modelred1, modelfull1) to see which one is better
anova(modelred1, modelfull1, test = "Chisq")
AIC(modelfull1, modelred1)
#although both test indicate the model with iteraction terms is better we end up choosing the additive model because it is easier to interpret and the auc for both the models for are similar. 
```

@tbl-summary organize the model results into an odds ratio table which proved the obesity prevalence, we consult the significance of obesity. Extremely low p-value that \<0.001 represents that all age group, full-time education and vegan,means these make a significant effect on obesity. The rest of predictors are not very significant.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summary
#| tbl-cap: Odds ratios from the logistic regression model for obesity prevalence
#| out.width: "80%"

sjPlot::tab_model(
  modelred1,
  transform = "exp",
  show.ci = F,
  show.p = TRUE,
  show.stat = TRUE,
  dv.labels = "Obesity (Yes vs No)"
)
```

Looking at the categories, male, vegetable, full-time education employment categories, paid employment, self-employment on gov't training, employment that works home have lower odds ratios. This means that they are less likely to be overweight. The other categories has higher odds and hence are more likely to be obese.

Age Group is statistically significant predictor of obesity with the risk increasing across age categories. In opposite, there is no significant difference between males and females. Furthermore, Most employment categories are also not statistically significant, although being a full-time education and consuming vegetables is associated with a lower risk of obesity.

Overall, the logistic regression results indicate that age is the strongest predictor of obesity. Lifestyle factors also play a role, as individuals who consume recommended amounts of vegetables are less likely to be obese. However, sex and most employment categories are not significantly associated with obesity.

Next, the performance of the final logistic regression model was assessed using the area under the ROC curve (AUC) in @fig-roc-auc. The model achieved an AUC of 0.59, suggesting that it performs slightly better than random guessing. This suggest that the model has a limited ability in distinguishing individuals who are obese and not obese. This is also reflected in ROC curve which shows that the curve remains relatively close to the line.

The low AUC also indicates that the explanatory variables included in the model do not sufficiently capture the factors associated with obesity prevalence. This may suggest that important predictors are missing from the dataset, such as genetics, physical activities, etc. Therefore, while the model identify some associations, its overall predictive performance is weak and should be interpreted with caution.

```{r}
#| echo: false
#| warning: false
#| message: false
#| results: hide
#Model fit with additive model 
pred1 <- ifelse(predict(modelred1, type = "response") > 0.5, "Yes", "No")

table(Predicted = pred1, Actual = data$Obese)
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-roc-auc
#| fig-cap: ROC curve and corresposding AUC of the logitsitc regression model
#| fig-align: center
#| fig-height: 3 
library(grid)
roc_obj1 <- roc(data$Obese, predict(modelred1, type = "response"))
plot(roc_obj1)
abline(a = 1, b = -1, lty = 2)
text(0.0, 0.6, paste("AUC =", round(auc(roc_obj1), 2)))
```

## Conclusion

This study examined obesity prevalence in Scotland using both exploratory analysis and logistic regression. The results suggest that obesity is influenced by multiple factors, with age identified as the most significant predictor, showing an increasing risk across older age groups. Dietary behaviour also plays a role, as individuals who meet recommended vegetable intake levels are less likely to be obese.

In contrast, variables such as sex, employment status, and year were not found to have statistically significant effects on obesity prevalence. While the model helps identify some meaningful associations, its overall predictive performance is relatively weak (AUC = 0.59), indicating limited ability to accurately classify individuals.

This suggests that important determinants of obesity—such as physical activity, genetic factors, and other lifestyle influences—may not be captured in the dataset. Therefore, although the analysis provides useful insights, obesity remains a complex issue that requires further investigation with more comprehensive data and improved modelling approaches.

<<<<<<< HEAD
## Limitations

-   **Omitted variables**: Key factors such as physical activity and genetic predisposition were not included, which may introduce bias.

-   **Limited predictive performance**: The model achieved a low AUC (0.59), indicating weak discriminatory ability.

-   **Cross-sectional design**: The data do not allow for causal inference, only associations can be identified.

-   **Self-reported bias**: Variables such as dietary intake may be subject to reporting inaccuracies.
=======
## Limitations 

-    **Omitted variables**: Key factors such as physical activity and genetic predisposition were not included, which may introduce bias.

-    **Limited predictive performance**: The model achieved a low AUC (0.59), indicating weak discriminatory ability.

-    **Cross-sectional design**: The data do not allow for causal inference, only associations can be identified.

-    **Self-reported bias**: Variables such as dietary intake may be subject to reporting inaccuracies.
>>>>>>> 2c36254b829f1457eee641edfec45aed92d65bad
