---
title: "Group 7 Data Analysis"
format: html
editor: visual
---

## Introduction

### Background

Obesity is a growing public health concern worldwide and is associated with an increased risk of various chronic conditions, including cardiovascular diseases, diabetes, and certain cancers. Monitoring the prevalence of obesity within a population is important for identifying trends over time and informing public health policies.

The Scottish Health Survey is designed to assess the health of individuals living in Scotland and to monitor changes in health-related behaviours and outcomes. The dataset used in this analysis contains information collected between 2008 and 2012, including demographic, socio-economic, and lifestyle factors alongside an indicator of obesity.

The primary aim of this study is to examine whether the prevalence of obesity in Scotland has changed over the survey years. In addition, the analysis will explore whether obesity prevalence differs across age groups, gender, employment status, and lifestyle behaviours such as fruit and vegetable consumption.

### Variables

The dataset contains the following variables:

-   **AgeGroup:** The age range of the individual, grouped in categories

-   **Sex:** The gender of the individual (Male/Female)

-   **Employment:** Employment status, used as an indicator of socio-economic status

-   **Fruit:** Whether the individual meets the recommended daily fruit intake (Yes/No)

-   **Veg:** Whether the individual meets the recommended daily vegetable intake (Yes/No)

-   **Year:** The year in which the survey was conducted (2008-2012)

-   **Obese:** Indicator variable showing whether the individual is classified as obese.

In this analysis, Obese is treated as the response variable. The remaining variables (age group, sex, employment status, and dietary factors) are considered explanatory variables to investigate potential differences in obesity prevalence across different population subgroups.

### Questions of interest

The main questions of interest are:

-   Has the prevalence of obesity in Scotland changed over the given years of the Scottish Health Survey?

-   Are there any differences in obesity by age, gender, socio-economic status or lifestyle factors?

## Exploratory data Analysis

```{r}
#| echo: false
#| warning: false
#| message: false
data<-read.csv("DAProject7.csv")
library(ggplot2)
library(tidyverse)
library(tidymodels)
library(gt)
library(MASS)
library(patchwork)
library(moderndive)
library(sjPlot)
library(performance)
library(kableExtra)
```

```{r}
#| echo: false
#| warning: false
#| message: false
table(data$Obese)
prop.table(table(data$Obese))
```

```{r}
#| echo: false
#| warning: false
#| message: false
ggplot(data, aes(x = Obese)) +
  geom_bar(fill = "skyblue") +
  theme_minimal() +
  labs(title = "Overall Obesity Prevalence",
       x = "Obese", y = "Count")
```

As the table and the figure shows. The proportion of individuals classified as obese provides an initial indication of obesity prevalence in the sample.

To explore whether obesity prevalence has changed over time, the proportion of individuals classified as obese was examined across different survey years.

```{r}
#| echo: false
#| warning: false
#| message: false
data %>%
  group_by(Year, Obese) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Year) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = factor(Year), y = prop, fill = Obese)) +
  geom_bar(stat = "identity", position = "fill") +
  theme_minimal() +
  labs(title = "Obesity Proportion by Year",
       x = "Year", y = "Proportion")
```

The proportion of obese individuals appears to vary slightly across years, suggesting potential changes in obesity prevalence over time.

To investigate whether obesity prevalence varies across different population subgroups, comparisons were made across demographic, socio-economic, and lifestyle factors.

```{r}
#| echo: false
#| warning: false
#| message: false
ggplot(data, aes(x = Sex, fill = Obese)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  labs(title = "Obesity by Sex",
       y = "Proportion")
```

There are slightly observable differences in obesity prevalence between males and females.

```{r}
#| echo: false
#| warning: false
#| message: false
ggplot(data, aes(x = AgeGroup, fill = Obese)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  labs(title = "Obesity by Age Group") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Obesity prevalence appears to differ across age groups.

```{r}
#| echo: false
#| warning: false
#| message: false
ggplot(data, aes(x = Employment, fill = Obese)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  labs(title = "Obesity by Employment Status") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Differences in obesity prevalence across employment groups suggest a possible socio-economic effect.

```{r}
#| echo: false
#| warning: false
#| message: false
data %>%
  pivot_longer(cols = c(Veg, Fruit),
               names_to = "Diet",
               values_to = "Intake") %>%
  ggplot(aes(x = Intake, fill = Obese)) +
  geom_bar(position = "fill") +
  facet_wrap(~ Diet) +
  theme_minimal() +
  labs(title = "Obesity by Diet Factors",
       x = "Intake (Yes/No)",
       y = "Proportion")
```

Obesity prevalence was also explored across dietary factors, including fruit and vegetable intake. The distributions appear similar, although slight differences may be observed between individuals meeting and not meeting the recommended intake.

The exploratory analysis suggests that obesity prevalence varies across different groups. In particular, differences are observed across age groups, and socio-economic and vegetable intake factors. These patterns indicate that obesity is likely influenced by multiple variables. A formal statistical analysis will be conducted to further investigate these relationships and determine their significance.

## Formal Analysis

### Logistic regression model

To investigate whether obesity prevalence in the Scottish population varies with factors such as age, sex, socioeconomic status, and lifestyle, we fitted a binary logistic regression model. Since the response variable "obesity" is a binary variable, logistic regression is suitable for simulating the probability of an individual being classified as obese.

$$
Y_i =
\begin{cases}
1, & \text{if individual } i \text{ is obese} \\
0, & \text{otherwise}
\end{cases}
$$

Let $Y_i = 1$ if individual $i$ is obese and $Y_i = 0$ otherwise. Then rewrite $Y_i$ as a binary logistic regression model

$$
Y_i \sim \text{Bernoulli}(p_i)
$$

The logistic regression model is written as

$$
\log\left(\frac{p_i}{1-p_i}\right)
=
\beta_0 + \beta_1 \text{AgeGroup} + \beta_2 \text{Sex} + \beta_3 \text{Employment}
+ \beta_4 \text{Veg} + \beta_5 \text{Fruit}.
$$

Where

-   $p_i$ the probability that individual $i$ is obese.
-   $\beta_0$ is the intercept of the model, corresponding to the baseline group.
-   $\beta_1{\text{AgeGroup}}$ is the differences in log-odds of obesity between each age category and the reference age group.
-   $\beta_2{\text{Sex}}$ is the difference in log-odds of obesity between males and females.
-   $\beta_3{\text{Employment}}$ is the differences in log-odds of obesity between each employment category and the reference employment group.
-   $\beta_4{\text{Veg}}$ is the difference in log-odds of obesity between individuals who do and do not meet the recommended vegetable intake.
-   $\beta_5{\text{Fruit}}$ is the difference in log-odds of obesity between individuals who do and do not meet the recommended fruit intake.

This model allows us to assess whether obesity prevalence varies with age group, sex, employment status, and dietary factors after adjusting for other explanatory variables.

### Model results

```{r}
#| echo: false
#| warning: false
#| message: false

data$Obese <- factor(data$Obese)
data$AgeGroup <- factor(data$AgeGroup)
data$Sex <- factor(data$Sex)
data$Employment <- factor(data$Employment)
data$Veg <- factor(data$Veg)
data$Fruit <- factor(data$Fruit)

Question2 <- glm(
  Obese ~ AgeGroup + Sex + Employment + Veg + Fruit,
  data = data,
  family = binomial)

summary(Question2)
```

@tbl-summary organize the model results into an odds ratio table which proved the obesity prevalence, we consult the significance of obesity.Extremely low p-value that \<0.001 represents that all age group, full-time education and vegan,means these make a significant effect on obesity. The rest of predictors are not very significant.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summary
#| tbl-cap: Odds ratios from the logistic regression model for obesity prevalence

sjPlot::tab_model(
  Question2,
  transform = "exp",
  show.ci = TRUE,
  show.p = TRUE,
  show.stat = TRUE,
  dv.labels = "Obesity (Yes vs No)"
)
```

For male,full-time education employment,paid ememployment, self-employedor on gov't training,employment that works home,vegetable and fruit consumer, they usually have lower odds ratios ,in a high obesity oddswhich means they are less likely to be overweight.The others are higher that more likely to be obese.

Obviously, 'Age Group' is statistically significant predictor. In opposite,there is no significant difference between males and females. Furthermore,being a full-time education and consuming vegetables is associated with a lower risk of obesity. The other employment and fruit consumption are not clearly indicates obesity.

### Analysis result

The logistic regression results indicate that age is the strongest predictor of obesity, with the risk increasing significantly across age groups. Lifestyle factors also play a role, as individuals who consume recommended amounts of vegetables are less likely to be obese. In contrast, sex and most employment categories are not significantly associated with obesity. That is all the related results we found by this project.
