---
title: "Group 7 Data Analysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Introduction

### Background

Obesity is a growing public health concern worldwide and is associated with an increased risk of various chronic conditions, including cardiovascular diseases, diabetes, and certain cancers. Monitoring the prevalence of obesity within a population is important for identifying trends over time and informing public health policies.

The Scottish Health Survey is designed to assess the health of individuals living in Scotland and to monitor changes in health-related behaviours and outcomes. The dataset used in this analysis contains information collected between 2008 and 2012, including demographic, socio-economic, and lifestyle factors alongside an indicator of obesity.

The primary aim of this study is to examine whether the prevalence of obesity in Scotland has changed over the survey years. In addition, the analysis will explore whether obesity prevalence differs across age groups, gender, employment status, and lifestyle behaviours such as fruit and vegetable consumption.

### Variables

The dataset contains the following variables:

-   **AgeGroup:** The age range of the individual, grouped in categories

-   **Sex:** The gender of the individual (Male/Female)

-   **Employment:** Employment status, used as an indicator of socio-economic status

-   **Fruit:** Whether the individual meets the recommended daily fruit intake (Yes/No)

-   **Veg:** Whether the individual meets the recommended daily vegetable intake (Yes/No)

-   **Year:** The year in which the survey was conducted (2008-2012)

-   **Obese:** Indicator variable showing whether the individual is classified as obese.

In this analysis, Obese is treated as the response variable. The remaining variables (age group, sex, employment status, and dietary factors) are considered explanatory variables to investigate potential differences in obesity prevalence across different population subgroups.

### Questions of interest

The main questions of interest are:

-   Has the prevalence of obesity in Scotland changed over the given years of the Scottish Health Survey?

-   Are there any differences in obesity by age, gender, socio-economic status or lifestyle factors?

## Exploratory data Analysis

```{r}
#| echo: false
#| warning: false
#| message: false
data<-read.csv("DAProject7.csv", stringsAsFactors = T)
str(data)
library(ggplot2)
library(tidyverse)
library(tidymodels)
library(gt)
library(MASS)
library(patchwork)
library(moderndive)
library(sjPlot)
library(performance)
library(kableExtra)
library(ggplot2)
library(broom)
library(yardstick)
library(pROC)

```

```{r}
#| echo: false
#| warning: false
#| message: false
table(data$Obese)
prop.table(table(data$Obese))
```

```{r}
#| echo: false
#| warning: false
#| message: false
ggplot(data, aes(x = Obese)) +
  geom_bar(fill = "skyblue") +
  theme_minimal() +
  labs(title = "Overall Obesity Prevalence",
       x = "Obese", y = "Count")
```

As the table and the figure shows. The proportion of individuals classified as obese provides an initial indication of obesity prevalence in the sample.

To explore whether obesity prevalence has changed over time, the proportion of individuals classified as obese was examined across different survey years.

```{r}
#| echo: false
#| warning: false
#| message: false
data %>%
  group_by(Year, Obese) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Year) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = factor(Year), y = prop, fill = Obese)) +
  geom_bar(stat = "identity", position = "fill") +
  theme_minimal() +
  labs(title = "Obesity Proportion by Year",
       x = "Year", y = "Proportion")
```

The proportion of obese individuals appears to vary slightly across years, suggesting potential changes in obesity prevalence over time.

To investigate whether obesity prevalence varies across different population subgroups, comparisons were made across demographic, socio-economic, and lifestyle factors.

```{r}
#| echo: false
#| warning: false
#| message: false
ggplot(data, aes(x = Sex, fill = Obese)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  labs(title = "Obesity by Sex",
       y = "Proportion")
```

There are slightly observable differences in obesity prevalence between males and females.

```{r}
#| echo: false
#| warning: false
#| message: false
ggplot(data, aes(x = AgeGroup, fill = Obese)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  labs(title = "Obesity by Age Group") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Obesity prevalence appears to differ across age groups.

```{r}
#| echo: false
#| warning: false
#| message: false
ggplot(data, aes(x = Employment, fill = Obese)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  labs(title = "Obesity by Employment Status") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Differences in obesity prevalence across employment groups suggest a possible socio-economic effect.

```{r}
#| echo: false
#| warning: false
#| message: false
data %>%
  pivot_longer(cols = c(Veg, Fruit),
               names_to = "Diet",
               values_to = "Intake") %>%
  ggplot(aes(x = Intake, fill = Obese)) +
  geom_bar(position = "fill") +
  facet_wrap(~ Diet) +
  theme_minimal() +
  labs(title = "Obesity by Diet Factors",
       x = "Intake (Yes/No)",
       y = "Proportion")
```

Obesity prevalence was also explored across dietary factors, including fruit and vegetable intake. The distributions appear similar, although slight differences may be observed between individuals meeting and not meeting the recommended intake.

The exploratory analysis suggests that obesity prevalence varies across different groups. In particular, differences are observed across age groups, and socio-economic and vegetable intake factors. These patterns indicate that obesity is likely influenced by multiple variables. A formal statistical analysis will be conducted to further investigate these relationships and determine their significance.

## Formal Analysis

```{r}
#| echo: false
#| warning: false
#| message: false

#testing the binomial model with just the additive terms.
modelred <- glm(Obese~ Year + Sex + AgeGroup + Employment + Veg + Fruit, data = data, family = binomial)
summary(modelred)

#did backward elimination using step funciton to get the final model as:
stats::step(modelred, direction = "backward")
modelred1 <- glm (Obese ~ Sex + AgeGroup + Employment + Veg, data = data, family = binomial)

#Testing model with a few chosen interaction terms 
modelfull <-  glm(Obese ~ Year + Sex + AgeGroup + Employment + Veg + Fruit +
            Year:Sex + Year:AgeGroup + Sex:AgeGroup,
          data = data, family = binomial)
summary(modelfull)

#backward elimintation using step function again
stats::step(modelfull, direction = "backward")

# got to the final model for this as:
modelfull1 <-  glm(Obese ~ Sex + AgeGroup + Employment + Veg + Sex:AgeGroup,
                    data = data, family = binomial)

#tested the both model (modelred1, modelfull1) to see which one is better
anova(modelred1, modelfull1, test = "Chisq")
AIC(modelfull1, modelred1)

#model fit of full model
pred <- ifelse(predict(model_full1, type = "response") > 0.5, "Yes", "No")

table(Predicted = pred, Actual = data$Obese)

roc_obj <- roc(data$Obese, predict(model_full1, type = "response"))
plot(roc_obj)
auc(roc_obj)
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: false

#Model fit with additive model 
pred1 <- ifelse(predict(modelred1, type = "response") > 0.5, "Yes", "No")

table(Predicted = pred1, Actual = data$Obese)

roc_obj1 <- roc(data$Obese, predict(modelred1, type = "response"))
plot(roc_obj1)
auc(roc_obj1)

#in terms of the AUC, the final model with interaction terms is better. And thus its also the reason why we choose the final model with the interaction terms
```
