---
title: "Group 7 Data Analysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Introduction

### Background

Obesity is a growing public health concern worldwide and is associated with an increased risk of various chronic conditions, including cardiovascular diseases, diabetes, and certain cancers. Monitoring the prevalence of obesity within a population is important for identifying trends over time and informing public health policies.

The Scottish Health Survey is designed to assess the health of individuals living in Scotland and to monitor changes in health-related behaviours and outcomes. The dataset used in this analysis contains information collected between 2008 and 2012, including demographic, socio-economic, and lifestyle factors alongside an indicator of obesity.

The primary aim of this study is to examine whether the prevalence of obesity in Scotland has changed over the survey years. In addition, the analysis will explore whether obesity prevalence differs across age groups, gender, employment status, and lifestyle behaviours such as fruit and vegetable consumption.

### Variables

The dataset contains the following variables:

-   **AgeGroup:** The age range of the individual, grouped in categories

-   **Sex:** The gender of the individual (Male/Female)

-   **Employment:** Employment status, used as an indicator of socio-economic status

-   **Fruit:** Whether the individual meets the recommended daily fruit intake (Yes/No)

-   **Veg:** Whether the individual meets the recommended daily vegetable intake (Yes/No)

-   **Year:** The year in which the survey was conducted (2008-2012)

-   **Obese:** Indicator variable showing whether the individual is classified as obese.

In this analysis, Obese is treated as the response variable. The remaining variables (age group, sex, employment status, and dietary factors) are considered explanatory variables to investigate potential differences in obesity prevalence across different population subgroups.

### Questions of interest

The main questions of interest are:

-   Has the prevalence of obesity in Scotland changed over the given years of the Scottish Health Survey?

-   Are there any differences in obesity by age, gender, socio-economic status or lifestyle factors?

## Exploratory data Analysis

```{r}
#| echo: false
#| warning: false
#| message: false
data<-read.csv("DAProject7.csv", stringsAsFactors = T)
str(data)
library(ggplot2)
library(tidyverse)
library(tidymodels)
library(gt)
library(MASS)
library(patchwork)
library(moderndive)
library(sjPlot)
library(performance)
library(kableExtra)


```

```{r}
#| echo: false
#| warning: false
#| message: false
table(data$Obese)
prop.table(table(data$Obese))
```

```{r}
#| echo: false
#| warning: false
#| message: false
ggplot(data, aes(x = Obese)) +
  geom_bar(fill = "skyblue") +
  theme_minimal() +
  labs(title = "Overall Obesity Prevalence",
       x = "Obese", y = "Count")
```

As the table and the figure shows. The proportion of individuals classified as obese provides an initial indication of obesity prevalence in the sample.

To explore whether obesity prevalence has changed over time, the proportion of individuals classified as obese was examined across different survey years.

```{r}
#| echo: false
#| warning: false
#| message: false
data %>%
  group_by(Year, Obese) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Year) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = factor(Year), y = prop, fill = Obese)) +
  geom_bar(stat = "identity", position = "fill") +
  theme_minimal() +
  labs(title = "Obesity Proportion by Year",
       x = "Year", y = "Proportion")
```

The proportion of obese individuals appears to vary slightly across years, suggesting potential changes in obesity prevalence over time.

To investigate whether obesity prevalence varies across different population subgroups, comparisons were made across demographic, socio-economic, and lifestyle factors.

```{r}
#| echo: false
#| warning: false
#| message: false
ggplot(data, aes(x = Sex, fill = Obese)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  labs(title = "Obesity by Sex",
       y = "Proportion")
```

There are slightly observable differences in obesity prevalence between males and females.

```{r}
#| echo: false
#| warning: false
#| message: false
ggplot(data, aes(x = AgeGroup, fill = Obese)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  labs(title = "Obesity by Age Group") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Obesity prevalence appears to differ across age groups.

```{r}
#| echo: false
#| warning: false
#| message: false
ggplot(data, aes(x = Employment, fill = Obese)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  labs(title = "Obesity by Employment Status") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Differences in obesity prevalence across employment groups suggest a possible socio-economic effect.

```{r}
#| echo: false
#| warning: false
#| message: false
data %>%
  pivot_longer(cols = c(Veg, Fruit),
               names_to = "Diet",
               values_to = "Intake") %>%
  ggplot(aes(x = Intake, fill = Obese)) +
  geom_bar(position = "fill") +
  facet_wrap(~ Diet) +
  theme_minimal() +
  labs(title = "Obesity by Diet Factors",
       x = "Intake (Yes/No)",
       y = "Proportion")
```

Obesity prevalence was also explored across dietary factors, including fruit and vegetable intake. The distributions appear similar, although slight differences may be observed between individuals meeting and not meeting the recommended intake.

The exploratory analysis suggests that obesity prevalence varies across different groups. In particular, differences are observed across age groups, and socio-economic and vegetable intake factors. These patterns indicate that obesity is likely influenced by multiple variables. A formal statistical analysis will be conducted to further investigate these relationships and determine their significance.

## Formal Analysis

To address the two research questions, a series of logistic regression models were fitted, since the response variable Obese is binary.

First, a simple logistic regression model including only Year was fitted. This model was used to address Question 1 by assessing whether obesity prevalence changed across the survey years.

```{r}
#| echo: false
#| warning: false
#| message: false

#testing with binomial model because of binary response with year for question 1
model_year <- glm(Obese~Year, data = data, family = binomial)
summary(model_year)

```

Next, to determine which model yielded better data results, we compared two models. Model 1 fitted an additive model that included year, gender, age group, employment status, vegetable intake, and fruit intake. This model included all factors, allowing us to test two research questions simultaneously.

```{r}
#| echo: false
#| warning: false
#| message: false

#testing the binomial model with just the additive terms.
modelred <- glm(Obese~ Year + Sex + AgeGroup + Employment + Veg + Fruit, data = data, family = binomial)
summary(modelred)

#did backward elimination using step funciton to get the final model as:
step(modelred, direction = "backward")
modelred1 <- glm (Obese ~ Sex + AgeGroup + Employment + Veg, data = data, family = binomial)
```

We used the `step()` function to perform backward elimination on the additive model. This resulted in a simplified model that included gender, age group, employment status, and vegetable intake, showing that year and fruit intake no longer had a significant effect after incorporating other variables were included.

By adding selected interaction terms (year:gender, year:age group, and gender:age group), we fitted a new interaction logistic regression model to explore whether the influence of one explanatory variable depends on another. After stepwise regression analysis, the final interaction model retained the interaction terms for gender, age group, employment status, vegetable intake, and gender and age group.

```{r}
#| echo: false
#| warning: false
#| message: false
#Testing model with a few chosen interaction terms 
modelfull <-  glm(Obese ~ Year + Sex + AgeGroup + Employment + Veg + Fruit +
            Year:Sex + Year:AgeGroup + Sex:AgeGroup,
          data = data, family = binomial)
summary(modelfull)

#backward elimintation using step function again
step(modelfull, direction = "backward")

# got to the final model for this as:
model_full1 <-  glm(Obese ~ Sex + AgeGroup + Employment + Veg + Sex:AgeGroup,
                    data = data, family = binomial)

#tested the both model (modelred1, modelfull1) to see which one is better
anova(model_full1, modelred1, test = "Chisq")
AIC(model_full1, modelred1)


```

The two candidate final models, Model1 and Model2, were compared using a likelihood ratio test and AIC. The interaction model Model2 was preferred because it provides a better representation of the data by allowing the relationship between sex and obesity to vary across age groups. Although it is slightly more complex, it captures an important interaction structure that the additive model does not. The ROC curve for Model2 also lies above the diagonal reference line, indicating some discriminatory ability.

```{r}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(sjPlot)
library(broom)
library(performance)
library(yardstick)
library(pROC)

#model fit of final model with interaction
pred <- ifelse(predict(model_full1, type = "response") > 0.5, "Yes", "No")

table(Predicted = pred, Actual = data$Obese)

roc_obj <- roc(data$Obese, predict(model_full1, type = "response"))
plot(roc_obj)
auc(roc_obj)
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: false

#Model fit with final additive model 
pred1 <- ifelse(predict(modelred1, type = "response") > 0.5, "Yes", "No")

table(Predicted = pred1, Actual = data$Obese)

roc_obj1 <- roc(data$Obese, predict(modelred1, type = "response"))
plot(roc_obj1)
auc(roc_obj1)
```

The final model therefore suggests that obesity prevalence is associated with age group, employment status, and vegetable intake, and that the effect of sex differs across age groups. In contrast, there is little evidence that obesity prevalence changed substantially across survey years once the other factors were taken into account. Similarly, fruit intake does not appear to have a strong independent association with obesity in the final model.
